Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-04-02 Ben_Manuscripts/hdphmm/Outline.tex
Calculate local density

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends, at least in part, on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  %MRS2: it can depend on other things, but ALSO depends on the latent state.
  %BJC2: added 'at least in part
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item The probability of transitioning to a state based on the current
    state is mathematically defined in terms of an $n\times n$ transition
    probability matrix, $T$, where $n$ is the number of states.
    \item Unfortunately, standard HMMs require $n$ to be known
    \textit{a priori}.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, H, a Dirichlet process 
    (DP) generates distributions over a countably infinite number of 
    probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the concentration 
    parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    %MRS3: This next line is a bit vauge (expresses confidence, closely related)
   \item The concentration parameter $\gamma$ expresses one's confidence in H relative to the posterior 
    and is closely related to the number of data observations.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

  \noindent We describe the dynamics of each state using a first order vector autoregressive
   (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
    %BJC2: The mean I'm talking about below is different than you are describing above.
    % The multivariate Gausian noise is coming from N(0, \Sigma)
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item We used a matrix-normal inverse-Wishart prior on parameters $A$ and $\Sigma$ 
  	and a Gaussian prior on $\mathbf{c}$ in order to infer their values.~\cite{fox_nonparametric_2009}
  	%BJC2: I'm not sure above is enough to make it clear how I got the parameters. Maybe I should
  	% explicitly refer to the fox paper above which outlines the inference procedure in more detail?
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden states
  while simultaneously estimating VAR(1) parameters for each state and the overall 
  state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %MRS3: be a bit more specific about what the criteria were for convergence.
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    as shown in SI.
    \item We refer the interested reader to much more extensive descriptions of 
    %MRS3: be more specific about which parts from the above are in the references
    this process and its implementation. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  %MRS3: be more specific - i.e. describe what are the ways you are going to apply it. 
  There are many ways one can apply the IHMM algorithm to timeseries data. 
  \begin{itemize}
    %MRS3: it's not as many as possible, but identify one's supported by the data, right?
    \item We begin by trying to identify as many distinct states as possible
    in each solute trajectory and then take advantage of the system's cylindrical
    symmetry in order to cluster the parameter sets into an interpretable number
    of states.
    %MRS3: not clear why this is the case?
    \item Having a large number of states gives more reliable clustering results.
%    \item We believe the subsequent procedure is best suited for our data
%    because it can clearly distinguish different dynamical behavior, it takes
%    advantage of the system's cylindrical symmetry and represents distinct
%    dynamics in an interpretable number of states.
  \end{itemize}

  We first applied the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    %MRS3: I'm not following the below? I don't know that we WANT to find more states, in my mind doing it this way 
    %the math is simpler this way since we don't have to rewrite things in term of radial diffusion
    %MRS3: we can brainstorm this as needed!
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

@@

  We applied the IHMM to each of the 24 solute trajectories independently.
  \begin{itemize}
    \item Although the IHMM is capable of identifying an infinite number of states, 
    a Dirichlet Process tends to exhibit a ``rich get richer" effect, favoring
    a fewer number of states.
    \item By applying the algorithm to each trajectory independently we reduce
    the possibility of lumping together multiple similar states which we
    would prefer to stay separated before clustering.
%    \item From a practical implementation standpoint, the algorithm is written
%    such that we must put a cap on the maximum number of states. We chose to 
%    allow up to 100 states for each trajectory although the algorithm typically
%    finds within the range of 5-15 states. 
    % BJC2: There will probably be a figure telling the number of states for 
    % each trajectory.
  \end{itemize}

  The states predicted by the IHMM are heavily influenced by the Gaussian prior
  placed on $c$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of the $A$ and $\mathbf{e}_t$ matrices do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to be halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the maximum and minimum
    to be 2 standard deviations 
    %MRS3: from the mean of what?
    from the mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    %MRS3: can you say more which categories converged quickly?
   \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We averaged the parameters from each iteration recorded after the
    equilibration time point.
    %MRS3: averaged and used this as estimate for running models later? 
    %MRS3: other possbility is maximum a posteriori estimate, but for peaked symmetric distributions, it's essentially the same.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    %MRS3: not sure what you mean that they all were multidimensional?
    \item Since all of the parameters are multidimensional, we found the 
    equilibration time point of each component and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because the mean in $r$ is 
  a useful clustering variable.
  %MRS3: because of the radial symmetry, x and y aren't really meaningful alone. 
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item In order to keep the same states found when we applied the algorithm
    in 3D, 
    %MRS3: would it be clearer to say you appplied only the inference part of the process on the clustered data?
    we reapplied the IHMM to the cylindrical trajectories with a fixed 
    state sequence, only allowing inference on the VAR(1) parameters. 
    %MRS3:did you cluster the state transitions as well? 
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  %MRS3: Maybe ``independent sets'' -> ``independent states, each with separate VAR(1) parameters?''
  \item For each solute studied, we identified between 200-300 independent sets of
   parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   %MRS3: group -> cluster
   to group them. % cite scikit-learn  
%   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on 5 variables: the radial means of each state, the two
  eigenvalues of $A$ and the two eigenvalues of $e_t$.
  \begin{itemize}
   \item Unfortunately, the number of independent parameter sets is too low for
   clustering in higher dimensions. 
   \item Ideally, we could use all of the entries of $A$ and $e_t$ in addition to 
   the radial means.
   \item It would not be helpful to use the means in the $z$ direction because solutes
   are essentially unbound in that direction, while 
   \item However, in our previous work, we observed radially dependent dynamical 
   behavior.~\cite{coscia_chemically_2019}  % cite new paper too.
  \end{itemize}

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states.
   %MRS3: seems like you should just be able to lump together the IHMM states to generate the IHMM estimate of T?
  \end{itemize}

@@

  %MRS3: ``used the IHMM algorithm'' -> ``used the parameter inference part of the IHMM algorithm''?
  We used the IHMM algorithm in order to infer $A$ and $\mathbf{e}_t$ of the 
  clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    %MRS3: not sure what is meant by ``Weighted by the rows of T''? Should it just be given by T? 
    \item We drew state sequences with transition probabilities weighted by the 
    the rows of $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}

>>>>

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends, at least in part, on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item The probability of transitioning to a state based on the current
    state is mathematically defined in terms of an $n\times n$ transition
    probability matrix, $T$, where $n$ is the number of states.
    \item Unfortunately, standard HMMs require $n$ to be known
    \textit{a priori}.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, $H$, a Dirichlet process 
    (DP) generates discrete distributions, $G_0$, over a countably infinite 
    number of probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the  
    concentration parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item Also expressed $G_0 \sim DP(\gamma, H)$
    %MRS3: This next line is a bit vauge (expresses confidence, closely related)
    %BJC3: tried to make more clear
    \item The concentration parameter, $\gamma$, expresses one's confidence in the 
    base distribution $H$. 
    \item When $\gamma\to 0$, the first weight of $G_0$, $\beta_1$, approaches 
    unity and for $\gamma\to\infty$, the weights become uniform and $G_0$ closely
    resembles $H$.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

  We describe the dynamics of each state visited by solutes using a first order vector 
  autoregressive (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
    %BJC2: The mean I'm talking about below is different than you are describing above.
    % The multivariate Gausian noise is coming from N(0, \Sigma)
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item This means that we only parameterize $A_1$. To simplify notation, we will just
  	call it $A$.
  	\item We used a matrix-normal inverse-Wishart prior on parameters $A$ and $\Sigma$ 
  	and a Gaussian prior on $\mathbf{c}$ in order to infer their values.~\cite{fox_nonparametric_2009}
  	%BJC2: I'm not sure above is enough to make it clear how I got the parameters. Maybe I should
  	% explicitly refer to the fox paper above which outlines the inference procedure in more detail?
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %BJC3: I got the software from this page: https://homes.cs.washington.edu/~ebfox/software/
    % Maybe I should reach out to emily fox to see how she would prefer I cite the software?
    %MRS3: be a bit more specific about what the criteria were for convergence.
    %BJC3: See if that's better
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    of each entry in the $c$ vector as well as the $A$ and $\Sigma$ matrices of 
    Equation~\ref{eqn:var}.
    \item Plots illustrating parameter convergence are available in the SI.
    \item We refer the interested reader to much more extensive descriptions of 
    %MRS3: be more specific about which parts from the above are in the references
    %BJC3: made more specific
    the inference and sampling procedures used to estimate the VAR(1) parameters
    and the state sequence. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  % BJC3: I tried again with this paragraph above. I want to set up the idea that we
  % are performing this somewhat complex procedure for a good reason before jumping
  % straight into it.
  %MRS3: be more specific - i.e. describe what are the ways you are going to apply it. 
%  There are many ways one can apply the IHMM algorithm to timeseries data. 
%  \begin{itemize}
%    %MRS3: it's not as many as possible, but identify one's supported by the data, right?
%    \item We begin by trying to identify as many distinct states as possible
%    in each solute trajectory and then take advantage of the system's cylindrical
%    symmetry in order to cluster the parameter sets into an interpretable number
%    of states.
%    %MRS3: not clear why this is the case?
%    \item Having a large number of states gives more reliable clustering results.
%%    \item We believe the subsequent procedure is best suited for our data
%%    because it can clearly distinguish different dynamical behavior, it takes
%%    advantage of the system's cylindrical symmetry and represents distinct
%%    dynamics in an interpretable number of states.
%  \end{itemize}

  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry.
  \begin{itemize}
%    \item In general we are interested in the time series created by tracing
%    each solute's center of mass position. % ALready said this above
    \item There are a number of viable ways in which one could choose to analyze 
    these time series.
    \item We could use all 3 dimensions straight from the output trajectory.
    \item We could use just the $z$ dimensional coordinate since we are primarily
    interested in through-plane membrane transport.
    \item We chose to use a multi-step procedure that we believe adequately
    distinguishes every type of distinct dynamical behavior exhibited by 
    the solutes.
    \item We work in cylindrical coordinates since each pore is cylindrical
    and we expect solutes to exhibit radially symmetric dynamics.
  \end{itemize}

  We start by applying the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    %MRS3: I'm not following the below? I don't know that we WANT to find more states, in my mind doing it this way 
    %the math is simpler this way since we don't have to rewrite things in term of radial diffusion
    %MRS3: we can brainstorm this as needed!
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

@@

  We applied the IHMM to each of the 24 solute trajectories independently.
  \begin{itemize}
    \item Although the IHMM is capable of identifying an infinite number of states, 
    a Dirichlet Process tends to exhibit a ``rich get richer" effect, favoring
    a fewer number of states.
    \item By applying the algorithm to each trajectory independently, we reduce
    the possibility of lumping together multiple similar states which we
    would prefer to stay separated before clustering.
%    \item From a practical implementation standpoint, the algorithm is written
%    such that we must put a cap on the maximum number of states. We chose to 
%    allow up to 100 states for each trajectory although the algorithm typically
%    finds within the range of 5-15 states. 
    % BJC2: There will probably be a figure telling the number of states for 
    % each trajectory.
  \end{itemize}

  The states identified by the IHMM are heavily influenced by the Gaussian prior
  placed on $\mathbf{c}$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of $A$ and $\Sigma$ do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to lie halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the
    maximum and minimum to be 2 standard deviations from the prior mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    %MRS3: can you say more which categories converged quickly?
    %BJC3: I imagine states with less sampling converge slower. But I'll need to
    % look further into this.
   \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We defined the finalized parameters of each state as the average
    of the parameters from each iteration recorded after the equilibration 
    time point.
    % MRS3: averaged and used this as estimate for running models later? 
    % BJC3: yes, well for clustering. Tried to clarified above
    % MRS3: other possbility is maximum a posteriori estimate, but for peaked 
    % symmetric distributions, it's essentially the same.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item We found the equilibration time point of each component of the 
    $\mathbf{c}$ vector as well as $A$ and $\Sigma$ matrices, and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because $x$ and $y$ aren't 
  meaningful alone due to the system's radial symmetry.
  % the mean in $r$ is a useful clustering variable.
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
%    \item In order to keep the same states found when we applied the algorithm
%    in 3D, we reapplied the IHMM to the cylindrical trajectories with a fixed 
%    state sequence, only allowing inference on the VAR(1) parameters.
    %MRS3: would it be clearer to say you appplied only the inference part of the process on the clustered data?
    %BJC3: like this?
    \item We fixed the state sequence from the 3D parmeterization and applied the
    inference component of the IHMM procedure to the cylindrical trajectories in
    order to estimate the VAR(1) parameters in terms of $r$ and $z$.
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  \item For each solute studied, we identified 200-300 independent states, each
  with separate VAR(1) parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   to cluster them. % cite scikit-learn  
%   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on 5 variables: the radial means of each state, the two
  eigenvalues of $A$ and the two eigenvalues of $\Sigma$.
  \begin{itemize}
   \item Unfortunately, the number of independent parameter sets is too low for
   clustering in higher dimensions. 
   \item Ideally, we could use all of the entries of $A$ and $\Sigma$ in addition to 
   the radial means.
   \item It would not be helpful to use the means in the $z$ direction because solutes
   are essentially unbound in that direction.
   \item However, in our previous work, we observed radially dependent dynamical 
   behavior.~\cite{coscia_chemically_2019}  % cite new paper eventually too.
  \end{itemize}

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states compiled across all 
   trajectories.
   %MRS3: seems like you should just be able to lump together the IHMM states to generate the IHMM estimate of T?
   %BJC3: That's what I do. Maybe I'm misunderstanding your suggestion. I added a little bit to the sentence
   % above to try and make more clear.
  \end{itemize}

@@

  %MRS3: ``used the IHMM algorithm'' -> ``used the parameter inference part of the IHMM algorithm''?
  We used the parameter inference component of the IHMM algorithm in order to 
  infer $A$ and $\Sigma$ of the clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    \item We drew state sequences with transition probabilities given by $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}
